Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-10 intro.tex
words about 2 conjecture

We focus on real-valued, scalar, time-series data that appear
\emph{complicated}: i.e., not periodic, linear, constant, etc.  We
make no assumptions about the properties of the generating process:
whether it is linear, nonlinear, deterministic, stochastic,
stationary, non-stationary, etc.  The information in an observation
can be partitioned into two pieces: redundancy and entropy
generation\cite{crutchfield2003}.  Our approach exploits this
decomposition in order to assess how much predictive structure is
present in a signal---i.e., where it falls on the complexity spectrum
mentioned above.  We define \emph{complexity} as a particular
approximation of Kolmogorov-Sinai entropy\cite{KS-entropy}.  That is,
we view a random-walk time series (which exhibits high entropy) as
purely complex, whereas a periodic signal, which exhibits low entropy,
is on the low end of the complexity spectrum.  We argue that an
extension of \emph{permutation entropy} \cite{bandt2002per}---a method
for approximating the entropy through ordinal analysis---is an
effective way to assess the complexity of a given time series.
Permutation entropy is ideal for our purposes because it is known to
converge to the true entropy value. Other existing techniques either require specific knowledge of the generating process orproduce biased values of the entropy~\cite{bollt2001}.

To explore the relationship between complexity, predictive structure,
and predictability, we generate forecasts of a number of time-series
datasets using four different prediction methods, then compare the
accuracy of those predictions to the permutation entropy of the
associated signals.  This results in two primary findings:
\begin{enumerate}
\item The complexity of a noisy real-valued time series is
  quantifiable by permutation entropy and correlated with prediction
  accuracy of an ideal predictor [[we don't know that.  we only tried
      four methods.]]
\item The way information is generated and processed internally by a
  system plays a crucial role in the success of different forecasting
  schema.
\end{enumerate}
The forecast methods used here were chosen to span the space of
standard prediction strategies, but they do not cover that space
exhaustively.  Our goal here is an empirical assessment of the
relationship between predictability and complexity, not formal results
about a ``best'' predictor for a given time series.  From a
practitioner's standpoint, it would be useful to know {\sl a priori}
if a trace contains enough predictive structure to make it worth
spending the time and effort searching for a good forecast method.  It
would also be useful to know if a given method is inadequate---that
is, if one could do better and should keep searching.

@@

[[Following belongs in the conclusion.  Putting it here would be
    appropriate in a CS paper, but not in a physics journal, and not
    when we've just made the point that this is just a testbed.]] 
% paragraph to appease the theoretician in me
It is worth taking a moment to consider the theoretical possibility of
this task. We are not attempting to predict the state of the CPU at an
arbitrary point in the future --- this, at least with perfect
accuracy, would be tantamount to solving the halting problem. What we
are attempting is to predict aspects or functions of the running of
the CPU, e.g., processor efficiency and similar statistics. Prediction
of these quantities at some finite time in the future, even with
perfect accuracy, does not violate the Rice-Shapiro
theorem~\cite{hopcroft2007} which essentially states that any non-trivial
property of a program is uncomputable.

>>>>

We focus on real-valued, scalar, time-series data that appear
\emph{complicated}: i.e., not periodic, linear, constant, etc.  We
make no assumptions about the properties of the generating process:
whether it is linear, nonlinear, deterministic, stochastic,
stationary, non-stationary, etc.  The information in an observation
can be partitioned into two pieces: redundancy and entropy
generation\cite{crutchfield2003}.  Our approach exploits this
decomposition in order to assess how much predictive structure is
present in a signal---i.e., where it falls on the complexity spectrum
mentioned above.  We define \emph{complexity} as a particular
approximation of Kolmogorov-Sinai entropy\cite{KS-entropy}.  That is,
we view a random-walk time series (which exhibits high entropy) as
purely complex, whereas a periodic signal, which exhibits low entropy,
is on the low end of the complexity spectrum.  We argue that an
extension of \emph{permutation entropy} \cite{bandt2002per}---a method
for approximating the entropy through ordinal analysis---is an
effective way to assess the complexity of a given time series.
Permutation entropy is ideal for our purposes because it is known to
converge to the true entropy value, unlike other existing techniques,
which either require specific knowledge of the generating process or
produce biased values of the entropy~\cite{bollt2001}.

To explore the relationship between complexity, predictive structure,
and predictability, we generate forecasts of a number of time-series
datasets using four different prediction methods, then compare the
accuracy of those predictions to the permutation entropy of the
associated signals.  This results in two primary findings:
\begin{enumerate}
\item The complexity of a noisy real-valued time series is
  quantifiable by permutation entropy and correlated with prediction
  accuracy of an ideal predictor [[we don't know that.  we only tried
      four methods.]]
\item The way information is generated and processed internally by a
  system plays a crucial role in the success of different forecasting
  schema.
\end{enumerate}

The second conjecture isn't just the corollary of halting problem in
disguise.  It gives insight into why there's not an ideal predictor.
It can't exist because the info is being transmitted in different
ways.

The forecast methods used here were chosen to span the space of
standard prediction strategies, but they do not cover that space
exhaustively.  Our goal here is an empirical assessment of the
relationship between predictability and complexity, not formal results
about a ``best'' predictor for a given time series.  From a
practitioner's standpoint, it would be useful to know {\sl a priori}
if a trace contains enough predictive structure to make it worth
spending the time and effort searching for a good forecast method.  It
would also be useful to know if a given method is inadequate---that
is, if one could do better and should keep searching.

@@

[[Following belongs in the conclusion.  Putting it here would be
    appropriate in a CS paper, but not in a physics journal, and not
    when we've just made the point that this is just a testbed.]] 
% paragraph to appease the theoretician in me
It is worth taking a moment to consider the theoretical possibility of
this task. We are not attempting to predict the state of the CPU at an
arbitrary point in the future --- this, at least with perfect
accuracy, would be tantamount to solving the halting problem. What we
are attempting is to predict aspects or functions of the running of
the CPU, e.g., processor efficiency and similar statistics. Prediction
of these quantities at some finite time in the future, even with
perfect accuracy, does not violate the Rice-Shapiro
theorem~\cite{hopcroft} which essentially states that any non-trivial
property of a program is uncomputable.
